---
title: "ReSpectR vs. Python: Direct Output Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ReSpectR vs. Python: Direct Output Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

find_pkg_root <- function() {
  candidates <- c(".", "..")
  for (p in candidates) {
    abs_p <- normalizePath(p, mustWork = FALSE)
    if (file.exists(file.path(abs_p, "DESCRIPTION")) && dir.exists(file.path(abs_p, "R"))) {
      return(abs_p)
    }
  }
  NULL
}

pkg_root <- find_pkg_root()
if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(ReSpectR)
}

find_extdata_path <- function() {
  has_required_assets <- function(p) {
    if (!nzchar(p) || !dir.exists(p)) {
      return(FALSE)
    }

    dir.exists(file.path(p, "time_tests")) &&
      dir.exists(file.path(p, "freq_tests")) &&
      dir.exists(file.path(p, "py_output_time")) &&
      dir.exists(file.path(p, "py_output_freq"))
  }

  pkg_path <- system.file("extdata", package = "ReSpectR")
  if (has_required_assets(pkg_path)) {
    return(pkg_path)
  }

  candidates <- c(
    file.path("inst", "extdata"),
    file.path("..", "inst", "extdata"),
    file.path("..", "..", "inst", "extdata")
  )

  for (p in candidates) {
    abs_p <- normalizePath(p, mustWork = FALSE)
    if (has_required_assets(abs_p)) {
      return(abs_p)
    }
  }

  stop(
    paste(
      "Could not locate required comparison assets.",
      "Expected time_tests/, freq_tests/, py_output_time/, and py_output_freq/ under installed package extdata or local inst/extdata."
    ),
    call. = FALSE
  )
}

base <- find_extdata_path()

# ------------------------------------------------------------------
# Helpers
# ------------------------------------------------------------------

# Run ReSpectR for one test, return crs and drs objects.
run_r_test <- function(domain, file_path, plateau = FALSE) {
  par <- setParams(
    domain = domain,
    dataFile = file_path,
    verbose = FALSE,
    plotting = FALSE,
    plateau = plateau
  )
  crs <- getContinuousSpectrum(par, writeOutput = FALSE)
  drs <- getDiscreteSpectrum(par, crs = crs, writeOutput = FALSE)
  list(crs = crs, drs = drs)
}

# Read Python precomputed output for one test.
# Returns a list with H (data.frame), Gfit (data.frame), dmodes (data.frame).
read_py_output <- function(domain, test_name, base) {
  sub <- if (domain == "time") "py_output_time" else "py_output_freq"
  d   <- file.path(base, sub, test_name)
  list(
    H      = utils::read.table(file.path(d, "H.dat"),      comment.char = "#"),
    Gfit   = utils::read.table(file.path(d, "Gfit.dat"),   comment.char = "#"),
    dmodes = utils::read.table(file.path(d, "dmodes.dat"), comment.char = "#")
  )
}

# Compare H(tau) spectrum: returns per-point abs difference and summary stats.
# Both Python and R use the same tau grid (identical algorithm, same ns, same data).
# Time domain H.dat: (tau, H)    — 2 columns.
# Freq domain H.dat: (tau, H, 0) — 3 columns (3rd always 0 for Im part).
compare_H <- function(r_s, r_H, py_H) {
  tau     <- r_s           # tau grid from R (== py_H$V1 by construction)
  H_R     <- r_H
  H_PY    <- py_H$V2
  diff    <- H_R - H_PY
  list(
    tau      = tau,
    H_R      = H_R,
    H_PY     = H_PY,
    diff     = diff,
    rmse     = sqrt(mean(diff^2)),
    max_diff = max(abs(diff))
  )
}

# Compare fitted G curves.
# Time:  Gfit columns (t, G_fit)           → compare 2nd column.
# Freq:  Gfit columns (w, G'_fit, G''_fit) → RMSE over both components.
compare_Gfit <- function(domain, r_Gfit, py_Gfit) {
  if (domain == "time") {
    diff <- r_Gfit[, 2] - py_Gfit$V2
    rmse <- sqrt(mean(diff^2))
    rel  <- sqrt(mean((diff / py_Gfit$V2)^2))
  } else {
    diff_p  <- r_Gfit[, 2] - py_Gfit$V2
    diff_pp <- r_Gfit[, 3] - py_Gfit$V3
    rmse    <- sqrt(mean(diff_p^2 + diff_pp^2))
    denom   <- sqrt(py_Gfit$V2^2 + py_Gfit$V3^2)
    rel     <- sqrt(mean((diff_p^2 + diff_pp^2) / denom^2))
  }
  list(rmse = rmse, rel_rmse = rel)
}

# Build a side-by-side comparison table of discrete modes.
# Pads the shorter mode list with NA so R and Python columns stay aligned.
side_by_side_modes <- function(r_modes, py_modes) {
  n <- max(nrow(r_modes), nrow(py_modes))
  pad <- function(x) {
    if (nrow(x) < n) {
      x[(nrow(x) + 1):n, ] <- NA
    }
    x
  }
  r2  <- pad(r_modes)
  py2 <- pad(py_modes)
  data.frame(
    tau_R  = r2$tau,
    g_R    = r2$g,
    tau_PY = py2$tau,
    g_PY   = py2$g
  )
}

# Plateau tests (frequency domain only).
plateau_tests <- c("test1u", "test7")

# Summary CSVs (still used for Python lamC reference scalar).
ref_H_time <- utils::read.csv(file.path(base, "comparison_H_time.csv"))
ref_H_freq <- utils::read.csv(file.path(base, "comparison_H_freq.csv"))
ref_D_time <- utils::read.csv(file.path(base, "comparison_disc_time.csv"))
ref_D_freq <- utils::read.csv(file.path(base, "comparison_disc_freq.csv"))
```

## Overview

This vignette computes fresh ReSpectR results **in-memory** (no file writes) and
compares them directly against precomputed Python (`pyReSpect`) output files
bundled in `inst/extdata`.

Three levels of comparison are performed per test:

1. **H(τ) spectrum** — point-by-point agreement of the continuous spectrum on the
   shared tau grid (identical algorithm ⟹ identical grid points).
2. **Fitted G values** — point-by-point agreement of G(t) (time) or G′/G″(ω)
   (frequency) produced by the continuous spectrum kernel reconstruction.
3. **Discrete modes** — comparison of Nopt and per-mode (τ, g) parameters.

Python reference files are in `inst/extdata/py_output_time/` and
`inst/extdata/py_output_freq/`, one sub-directory per test case.

## Time-Domain Comparison

```{r time-comparison}
time_tests <- paste0("test", 1:7)
time_dir   <- file.path(base, "time_tests")

time_results <- lapply(time_tests, function(tt) {
  dat_file  <- file.path(time_dir, paste0(tt, ".dat"))
  stopifnot(file.exists(dat_file))

  r   <- run_r_test("time", dat_file)
  py  <- read_py_output("time", tt, base)

  hc    <- compare_H(r$crs$s, r$crs$H, py$H)
  gfc   <- compare_Gfit("time", r$crs$Gfit, py$Gfit)

  # Discrete modes: Python time dmodes = (g, tau, dtau) — 3 columns
  py_Nopt  <- nrow(py$dmodes)
  # R modes sorted by tau for comparison
  r_modes  <- data.frame(tau = r$drs$tau, g = r$drs$g)
  r_modes  <- r_modes[order(r_modes$tau), ]
  py_modes <- data.frame(tau = py$dmodes$V2, g = py$dmodes$V1)
  py_modes <- py_modes[order(py_modes$tau), ]

  # lamC from summary reference
  lam_ref <- ref_H_time[ref_H_time$test == paste0(tt, ".dat"), ]

  list(
    test         = tt,
    tau          = hc$tau,
    H_R          = hc$H_R,
    H_PY         = hc$H_PY,
    lamC_R       = r$crs$lamC,
    lamC_PY      = if (nrow(lam_ref) > 0) lam_ref$lamC_PY else NA,
    Nopt_R       = r$drs$Nopt,
    Nopt_PY_file = py_Nopt,
    H_rmse       = hc$rmse,
    H_max_diff   = hc$max_diff,
    Gfit_rmse    = gfc$rmse,
    Gfit_rel     = gfc$rel_rmse,
    r_modes      = r_modes,
    py_modes     = py_modes
  )
})
```

### H(τ) Spectrum Agreement (Time)

RMSE and max |ΔH| are on the H-value scale (the same log-space units used
internally by both solvers).

```{r time-H-table}
time_H_tbl <- do.call(rbind, lapply(time_results, function(r)
  data.frame(
    test       = r$test,
    lamC_R     = round(r$lamC_R,  4),
    lamC_PY    = round(r$lamC_PY, 4),
    H_rmse     = signif(r$H_rmse,     3),
    H_max_diff = signif(r$H_max_diff, 3)
  )))
time_H_tbl
```

### Fitted G(t) Agreement (Time)

```{r time-Gfit-table}
time_Gfit_tbl <- do.call(rbind, lapply(time_results, function(r)
  data.frame(
    test        = r$test,
    Gfit_rmse   = signif(r$Gfit_rmse, 3),
    Gfit_rel    = signif(r$Gfit_rel,  3)
  )))
time_Gfit_tbl
```

### Discrete Modes (Time)

`Nopt_R` is the count returned by `getDiscreteSpectrum()`.
`Nopt_PY_file` counts rows in the Python `dmodes.dat` — Python omits modes
that collapsed to the grid boundary during refinement, so this can be slightly
lower than the declared optimal Nopt.

```{r time-modes-table}
time_modes_tbl <- do.call(rbind, lapply(time_results, function(r)
  data.frame(
    test         = r$test,
    Nopt_R       = r$Nopt_R,
    Nopt_PY_file = r$Nopt_PY_file
  )))
time_modes_tbl
```

Per-mode detail (τ, g) for the first two time tests:

```{r time-modes-detail, results = "asis"}
for (i in 1:2) {
  res <- time_results[[i]]
  tbl <- side_by_side_modes(res$r_modes, res$py_modes)
  cat(sprintf("\n**%s — Side-by-side discrete modes:**\n\n", res$test))
  print(knitr::kable(round(tbl, 5), digits = 5,
                     col.names = c("tau_R", "g_R", "tau_PY", "g_PY")))
}
```

### H(τ) Overlay Plot (Time)

```{r time-H-plot, fig.width = 10, fig.height = 5}
op <- par(mfrow = c(2, 4), mar = c(3, 3, 2, 1))
for (res in time_results) {
  plot(log10(res$tau), res$H_R, type = "l",
    col = "#1b6ca8", lwd = 2,
    main = res$test, xlab = "log10(τ)", ylab = "H")
  lines(log10(res$tau), res$H_PY, col = "#c44e52", lwd = 1, lty = 2)
  legend("topleft", c("R", "Python"), col = c("#1b6ca8", "#c44e52"),
         lwd = c(2, 1), lty = c(1, 2), bty = "n", cex = 0.7)
}
par(op)
```

## Frequency-Domain Comparison

```{r freq-comparison}
freq_tests <- c("test1", "test1n", "test1u", "test2", "test3",
                "test4", "test5", "test6", "test7")
freq_dir   <- file.path(base, "freq_tests")

freq_results <- lapply(freq_tests, function(tt) {
  dat_file <- file.path(freq_dir, paste0(tt, ".dat"))
  stopifnot(file.exists(dat_file))

  r  <- run_r_test("frequency", dat_file, plateau = tt %in% plateau_tests)
  py <- read_py_output("frequency", tt, base)

  hc  <- compare_H(r$crs$s, r$crs$H, py$H)
  gfc <- compare_Gfit("frequency", r$crs$Gfit, py$Gfit)

  # Discrete modes: Python freq dmodes = (g, tau) — 2 columns
  py_Nopt  <- nrow(py$dmodes)
  r_modes  <- data.frame(tau = r$drs$tau, g = r$drs$g)
  r_modes  <- r_modes[order(r_modes$tau), ]
  py_modes <- data.frame(tau = py$dmodes$V2, g = py$dmodes$V1)
  py_modes <- py_modes[order(py_modes$tau), ]

  lam_ref <- ref_H_freq[ref_H_freq$test == paste0(tt, ".dat"), ]

  list(
    test         = tt,
    lamC_R       = r$crs$lamC,
    lamC_PY      = if (nrow(lam_ref) > 0) lam_ref$lamC_PY else NA,
    Nopt_R       = r$drs$Nopt,
    Nopt_PY_file = py_Nopt,
    H_rmse       = hc$rmse,
    H_max_diff   = hc$max_diff,
    Gfit_rmse    = gfc$rmse,
    Gfit_rel     = gfc$rel_rmse,
    r_modes      = r_modes,
    py_modes     = py_modes
  )
})
```

### H(τ) Spectrum Agreement (Frequency)

For the frequency domain the H vector stores the spectral amplitude on the same
shared τ grid; the third column of `H.dat` is zero (imaginary part is not used).

```{r freq-H-table}
freq_H_tbl <- do.call(rbind, lapply(freq_results, function(r)
  data.frame(
    test       = r$test,
    lamC_R     = round(r$lamC_R,  4),
    lamC_PY    = round(r$lamC_PY, 4),
    H_rmse     = signif(r$H_rmse,     3),
    H_max_diff = signif(r$H_max_diff, 3)
  )))
freq_H_tbl
```

### Fitted G*(ω) Agreement (Frequency)

```{r freq-Gfit-table}
freq_Gfit_tbl <- do.call(rbind, lapply(freq_results, function(r)
  data.frame(
    test      = r$test,
    Gfit_rmse = signif(r$Gfit_rmse, 3),
    Gfit_rel  = signif(r$Gfit_rel,  3)
  )))
freq_Gfit_tbl
```

### Discrete Modes (Frequency)

```{r freq-modes-table}
freq_modes_tbl <- do.call(rbind, lapply(freq_results, function(r)
  data.frame(
    test         = r$test,
    Nopt_R       = r$Nopt_R,
    Nopt_PY_file = r$Nopt_PY_file
  )))
freq_modes_tbl
```

Per-mode detail for test1 frequency:

```{r freq-modes-detail, results = "asis"}
res <- freq_results[[1]]
tbl <- side_by_side_modes(res$r_modes, res$py_modes)
cat("\n**test1 (frequency) — Side-by-side discrete modes:**\n\n")
print(knitr::kable(round(tbl, 5), digits = 5,
                   col.names = c("tau_R", "g_R", "tau_PY", "g_PY")))
```

## Aggregate Summary

```{r aggregate}
agg_time <- data.frame(
  domain         = "time",
  mean_H_rmse    = mean(sapply(time_results,  `[[`, "H_rmse")),
  max_H_rmse     = max(sapply(time_results,   `[[`, "H_rmse")),
  mean_Gfit_rmse = mean(sapply(time_results,  `[[`, "Gfit_rmse")),
  mean_Nopt_diff = mean(abs(sapply(time_results,  `[[`, "Nopt_R") -
                            sapply(time_results,  `[[`, "Nopt_PY_file")))
)
agg_freq <- data.frame(
  domain         = "frequency",
  mean_H_rmse    = mean(sapply(freq_results,  `[[`, "H_rmse")),
  max_H_rmse     = max(sapply(freq_results,   `[[`, "H_rmse")),
  mean_Gfit_rmse = mean(sapply(freq_results,  `[[`, "Gfit_rmse")),
  mean_Nopt_diff = mean(abs(sapply(freq_results,  `[[`, "Nopt_R") -
                            sapply(freq_results,  `[[`, "Nopt_PY_file")))
)
rbind(agg_time, agg_freq)
```

## Notes

- All R results are computed fresh in-memory (`writeOutput = FALSE`).
- Python output files (`H.dat`, `Gfit.dat`, `dmodes.dat`) were precomputed with
  the original `pyReSpect` code and are bundled in `inst/extdata/py_output_time/`
  and `inst/extdata/py_output_freq/`.
- Python and R use an identical tau grid formula: both generate `ns = 100` points
  geometrically spaced between the same `smin` and `smax` derived from the data,
  so H(τ) can be compared point-by-point without interpolation.
- `lamC_PY` values come from the packaged summary CSVs (`comparison_H_*.csv`)
  rather than from the individual output files, since the Python code selects
  the optimal λ during the L-curve search rather than writing it in a parseable
  scalar form.
- `Nopt_PY_file` counts rows in Python's `dmodes.dat`.  Python omits modes that
  collapsed to grid boundaries during optimisation, so this number can differ
  from R's `Nopt` without indicating a real algorithmic discrepancy.
