---
title: "Comparing ReSpectR Results to Python"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing ReSpectR Results to Python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)

find_pkg_root <- function() {
  candidates <- c(".", "..")
  for (p in candidates) {
    abs_p <- normalizePath(p, mustWork = FALSE)
    if (file.exists(file.path(abs_p, "DESCRIPTION")) && dir.exists(file.path(abs_p, "R"))) {
      return(abs_p)
    }
  }
  NULL
}

pkg_root <- find_pkg_root()
if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(ReSpectR)
}

# Resolve extdata both for installed package runs and local dev runs (RStudio).
find_extdata_path <- function() {
  has_required_assets <- function(p) {
    if (!nzchar(p) || !dir.exists(p)) {
      return(FALSE)
    }

    required_files <- c(
      "comparison_H_time.csv",
      "comparison_H_freq.csv",
      "comparison_disc_time.csv",
      "comparison_disc_freq.csv"
    )

    all(file.exists(file.path(p, required_files))) &&
      dir.exists(file.path(p, "time_tests")) &&
      dir.exists(file.path(p, "freq_tests"))
  }

  pkg_path <- system.file("extdata", package = "ReSpectR")
  if (has_required_assets(pkg_path)) {
    return(pkg_path)
  }

  candidates <- c(
    file.path("inst", "extdata"),
    file.path("..", "inst", "extdata"),
    file.path("..", "..", "inst", "extdata")
  )

  for (p in candidates) {
    abs_p <- normalizePath(p, mustWork = FALSE)
    if (has_required_assets(abs_p)) {
      return(abs_p)
    }
  }

  stop(
    paste(
      "Could not locate required comparison assets.",
      "Expected comparison_*.csv plus time_tests/ and freq_tests/ under installed package extdata or local inst/extdata."
    ),
    call. = FALSE
  )
}

base <- find_extdata_path()

run_single_test <- function(domain, file_path, plateau = FALSE) {
  par <- setParams(
    domain = domain,
    dataFile = file_path,
    verbose = FALSE,
    plotting = FALSE,
    plateau = plateau
  )

  crs <- getContinuousSpectrum(par, writeOutput = FALSE)
  drs <- getDiscreteSpectrum(par, crs = crs, writeOutput = FALSE)

  list(lamC = crs$lamC, Nopt = drs$Nopt)
}
```

## Overview

This vignette computes fresh ReSpectR results in-memory (no file writes) and
compares them against packaged Python reference summaries.

Reference tables are distributed with the package in `inst/extdata`.

## Available comparison files

```{r files}
list.files(base, pattern = "^comparison_.*\\.csv$", full.names = TRUE)
```

Expected files:

- `comparison_H_time.csv`
- `comparison_H_freq.csv`
- `comparison_disc_time.csv`
- `comparison_disc_freq.csv`

Expected bundled test-data directories:

- `time_tests/`
- `freq_tests/`

## Fresh Time-Domain Comparison

```{r time-fresh}
ref_H_time <- utils::read.csv(file.path(base, "comparison_H_time.csv"))
ref_D_time <- utils::read.csv(file.path(base, "comparison_disc_time.csv"))

time_tests <- sort(unique(ref_H_time$test))
time_dir <- file.path(base, "time_tests")

time_current <- lapply(time_tests, function(tt) {
  dat_file <- file.path(time_dir, paste0(tt, ".dat"))
  stopifnot(file.exists(dat_file))
  cur <- run_single_test("time", dat_file)
  data.frame(test = tt, lamC_current = cur$lamC, Nopt_current = cur$Nopt)
})
time_current <- do.call(rbind, time_current)

time_compare <- merge(time_current, ref_H_time[, c("test", "lamC_R", "lamC_PY")], by = "test")
time_compare <- merge(time_compare, ref_D_time[, c("test", "Nopt_R", "Nopt_PY")], by = "test")

time_compare$lamC_rel_diff_vs_PY <- abs(time_compare$lamC_current - time_compare$lamC_PY) /
  pmax(abs(time_compare$lamC_PY), .Machine$double.eps)
time_compare$lamC_rel_diff_vs_archived_R <- abs(time_compare$lamC_current - time_compare$lamC_R) /
  pmax(abs(time_compare$lamC_R), .Machine$double.eps)
time_compare$Nopt_delta_vs_PY <- time_compare$Nopt_current - time_compare$Nopt_PY
time_compare$Nopt_delta_vs_archived_R <- time_compare$Nopt_current - time_compare$Nopt_R

time_compare[order(time_compare$test), ]
```

## Fresh Frequency-Domain Comparison

```{r freq-fresh}
ref_H_freq <- utils::read.csv(file.path(base, "comparison_H_freq.csv"))
ref_D_freq <- utils::read.csv(file.path(base, "comparison_disc_freq.csv"))

freq_tests <- sort(unique(ref_H_freq$test))
freq_dir <- file.path(base, "freq_tests")

plateau_tests <- c("test1u", "test7")

freq_current <- lapply(freq_tests, function(tt) {
  dat_file <- file.path(freq_dir, paste0(tt, ".dat"))
  stopifnot(file.exists(dat_file))
  cur <- run_single_test("frequency", dat_file, plateau = tt %in% plateau_tests)
  data.frame(test = tt, lamC_current = cur$lamC, Nopt_current = cur$Nopt)
})
freq_current <- do.call(rbind, freq_current)

freq_compare <- merge(freq_current, ref_H_freq[, c("test", "lamC_R", "lamC_PY")], by = "test")
freq_compare <- merge(freq_compare, ref_D_freq[, c("test", "Nopt_R", "Nopt_PY")], by = "test")

freq_compare$lamC_rel_diff_vs_PY <- abs(freq_compare$lamC_current - freq_compare$lamC_PY) /
  pmax(abs(freq_compare$lamC_PY), .Machine$double.eps)
freq_compare$lamC_rel_diff_vs_archived_R <- abs(freq_compare$lamC_current - freq_compare$lamC_R) /
  pmax(abs(freq_compare$lamC_R), .Machine$double.eps)
freq_compare$Nopt_delta_vs_PY <- freq_compare$Nopt_current - freq_compare$Nopt_PY
freq_compare$Nopt_delta_vs_archived_R <- freq_compare$Nopt_current - freq_compare$Nopt_R

freq_compare[order(freq_compare$test), ]
```

## Aggregate Similarity Summary

```{r summary}
summary_stats <- data.frame(
  domain = c("time", "frequency"),
  mean_lamC_rel_diff_vs_PY = c(
    mean(time_compare$lamC_rel_diff_vs_PY, na.rm = TRUE),
    mean(freq_compare$lamC_rel_diff_vs_PY, na.rm = TRUE)
  ),
  max_lamC_rel_diff_vs_PY = c(
    max(time_compare$lamC_rel_diff_vs_PY, na.rm = TRUE),
    max(freq_compare$lamC_rel_diff_vs_PY, na.rm = TRUE)
  ),
  mean_abs_Nopt_delta_vs_PY = c(
    mean(abs(time_compare$Nopt_delta_vs_PY), na.rm = TRUE),
    mean(abs(freq_compare$Nopt_delta_vs_PY), na.rm = TRUE)
  )
)

summary_stats
```

## Notes

- This vignette does not write output files; all current ReSpectR results are
  computed in-memory via `writeOutput = FALSE`.
- Python raw output vectors are not distributed in the package, so fresh
  similarity is assessed against packaged Python reference summary metrics
  (`lamC_PY`, `Nopt_PY`) and drift versus archived R reference columns.
